Towards a Verified Bzip2

### 2019-10-08 draft_paper/draft_paper.tex
Finish first draf of new intro and abstract

\begin{abstract}
Text of abstract \ldots.
\end{abstract}

@@

\keywords{Formal methods, Sorting and seraching, Data compression}

@@

The Burrows--Wheeler transform (BWT) is an invertible transformation
that makes a string more amenable to compression by other methods
\cite{bw}. It accomplishes this by creating runs of repeated letters
in the produced string. Of course, a simple way to create runs is to
sort the string, but that would be useless for compression because the
original string would be unrecoverable.

>>>>

\begin{abstract}
Stable sorts are an important and fundamental type of sorting
algorithm. Some algorithms, such as radix sorts, rely upon the
stability of a sorting subroutine for their correctness. In other
cases, stable sorts are useful for reasoning about algorithms, often
because any two stable sorts are interchangeable. One such algorithm
is the Burrows--Wheeler transform (BWT), which has applications in
compression, full-text indexing and pattern searching.

There has not been much work on formally verifying stable sorts, or
rather, the stability of stable sorts. In this paper, we develop more
fully the theory of stable sorts, and use this theory to prove the
correctness of radixsort and the BWT.
\end{abstract}

@@

\keywords{Formal methods, Sorting and searching, Data compression}

@@

Stable sorts are an important and fundamental type of sorting
algorithm. A sort is stable if it preserves the relative order of
equal elements: for example, if we are sorting a list of words by the
first characters, every word that starts with \texttt{a} appears in
the same order relative to the other words that start with \texttt{a}
as in the original list. Stability can be directly useful in certain
contexts, such as when sorting a table by a single column. However,
stability is also crucial to correctness of certain general-purpose
algorithms.

One such algorith, or set of algorithms, are radix sorts. Radix sorts
operate on lists with fixed-size elements, such as strings of a given
length or integers with a given number of digits. A Least Significant
Digit (LSD) radix sort uses some stable sort to sort each column,
starting with the rightmost (or least significant) and ending with the
leftmost (or most significant) digit. Radix sorts can be faster than
other comparison-based sorts such as quicksort, achieving linear
worst-case runtimes in the length of the list. This is because they
are less general, in a sense: rather than relying only on a comparison
function, they exploit more of the elements' structure. The cost of
requiring that structure is that we are restricted to lists with
fixed-size elements. In practice, though, many common types of data
are of fixed size: integers, floating point numbers, and fixed-length
strings, as well as structs and tuples of other fixed size data types.
In addition, radix sorts can be extended to handle variable-length
strings \cite{McIlroy93}.

% https://probablydance.com/2016/12/02/investigating-radix-sort/

Stable sorts are also important for reasoning about sorting algorithms
more broadly. This is because any two stable sorting algorithms are
equivalent, in the sense that they will produce the same outputs for a
given input. Crucially, this allows us to reason using elegant
algorithms and replace them with efficient algorithms later on. We
make use of this technique in order to reason about LSD radix sort, as
well as an algorithm called the Burrows-Wheeler transform (BWT).

The Burrows--Wheeler transform (BWT) is an invertible transformation
that makes a string more amenable to compression by other methods
\cite{bw}. It accomplishes this by creating runs of repeated letters
in the produced string. Of course, a simple way to create runs is to
sort the string, but that would be useless for compression because the
original string would be unrecoverable. Notably, the BWT is the core
pass in the popular, open-source bzip2 compression software
\cite{tsai_2016}. In addition, the BWT has been recently applied to
full-text indexing and pattern searching, for example in genomics
\cite{ferragina_index, dna}.

There has not been much work on formally verifying stable sorts, or
rather, the stability of stable sorts. Some work has been done on
verifying the stability of mergesort
\cite{leroy_2018,Sternagel2013,Leino2015,deGouw2014}, but we do not
know of any formally verified implementation of a radix sort including
stability. In addition, to our knowledge, no one has produced a
formally verified implementation of the BWT, by which we mean an
implementation of the forwards-transform, the inverse-transform, and a
proof that they the inverse-transform inverts the forwards-transform.

In this paper, we develop more fully the theory of stable sorts: we
define stability in 3 different ways and prove equivalence between
them, and we prove that all stable sorts are interchangeable. We use
this to prove the correctness and stability of an LSD radix sort.
Finally, we prove implement and prove correct the BWT, basing our
proof on the excellent derivation of Bird
et~al\cite{alg-of-prog,birdmu}. Bird's technique of proof by program
calculation transferred very well to a machine-checked setting; the
most difficult part was proving the lemmas the derivation relies upon,
which include the correctness of LSD radix sort.

\section{Main Ideas}
